---
title: "SNAP Data 1990-2024"
format: 
    html:
      
      code-fold: true 
toc: true
  
theme: mycustom.scss
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE, message=FALSE, warning=FALSE}

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)
library(highcharter)
library(sjlabelled)

source("functionsg.R")

snap_final <- read_rds("data/snap_df3.rds")

# library(showtext)
# font_add_google(name = "Montserrat", family = "Montserrat")
# showtext_auto()


extrafont::loadfonts(quiet = FALSE)

mysource <- "Source: U.S. Department of Agriculture 2024"



fig_basic <- snap_final %>% 
  filter(state_abbrev == "US") %>% 
  rename("Number of households" = snaphh2, 
         date = ym) %>% 
  ggplot(aes(x=date, y= `Number of households`))+
  geom_col(alpha = 0.5, fill = nord_silver[5])+
  custom_style3()+
  scale_x_date(name = NULL)+
  scale_y_continuous(name = NULL, 
                   labels = scales::comma)+
  labs(title = "Number of US Households Receiving SNAP Benefits", 
       caption = mysource)

  
# plotly::ggplotly(fig_basic) %>% 
#   plotly::layout(font = list(family = "Montserrat"), 
#                  title = list(x = 0.4,            # x position with respect to `xref` in normalized 
#                                         # coordinates from "0" (left) to "1" (right).
#                       xanchor ="right"), 
#                  xaxis = list(title = ''), 
#                  yaxis = list(title = ''), 
#                   margin = list(
#       l = 0,
#       r = 0,
#       b = 0,
#       t = 40, pad = 1
#     ))

```

## Number of households receiving benefits

Supplemental Nutritional Assistance Program (SNAP) is a program from the US Department of Agriculture that provides nutritional assistance to low income households and individuals. In September 2024, more than 22.5 million households received SNAP benefits. The number of households receiving assistance has stayed above 21 million since April 2020, just below the record amount between 2012 and 2014.

```{r}

snap_final %>% 
  filter(state_abbrev == "US") %>% 
  hchart(., "column",
         hcaes(x = ym, y = round(snaphh2)),
         color = nord_silver[5], 
         name = "Number of households") %>% 
  hc_title(text = "Number of US Households Receiving SNAP Benefits") %>% 
  hc_xAxis(title = list(text = "")) %>% 
  hc_yAxis(title = list(text = "")) %>% 
  hc_caption(text = "Source: U.S. Department of Agriculture") %>% 
  hc_add_theme(hc_theme_cstm())

```

## Benefit amounts over time

Current benefits average around \$350 per household and about \$190 per person. Benefit amounts peaked during the more generous COVID relief period in Nov 2022 averaging \$260 per person and \$490 per household.

```{r}


snap_final %>% 
  filter(state_abbrev == "US") %>% 
  hchart(., "line",
         hcaes(x = ym, y= benperhh), 
         color = nord_silver[5], 
         name = "Benefit amount per household", 
         tooltip = list(valueDecimals = 0, valuePrefix = "$")) %>% 
  hc_add_series(data = snap_final %>%
                  filter(state_abbrev == "US"),
                "line", 
         hcaes(x = ym, y = round(benperp)),
         color = nord_silver[1],
         name = "Benefit amount per person", 
         tooltip = list(valueDecimals = 0, valuePrefix = "$")) %>% 
  hc_title(text = "Benefit Amounts per Household and per Person") %>%
  hc_xAxis(title = list(text = "")) %>% 
  hc_yAxis(title = list(text = "")) %>% 
  hc_caption(text = "Source: U.S. Department of Agriculture") %>% 
  hc_add_theme(hc_theme_cstm())









```

## Comparisons at the state and regional levels

::: column-body-outset
```{r, fig.width=10, fig.height=8}


fig2 <- snap_final %>% 
  filter(!is.na(region_name)) %>% 
  ggplot(aes(x = ym, y = pct_snapp3))+
  geom_line(aes(group = state_abbrev),
            alpha = 0.3, 
             col = Maizgray) +
  geom_smooth(method = "loess", col = nord_silver[5])+
  facet_wrap(~region_name, 
             nrow = 1)+
  custom_style3()+
  scale_y_continuous(name = NULL, 
                     labels = add_pct_sign)+
  scale_x_date(name = NULL)+
  theme(plot.background = element_rect(fill = "#F0F0F0"), 
        strip.background = element_rect(color =  "#bfbfbf",
                                        fill = "#bfbfbf"), 
        strip.text = element_text(family = "Montserrat",
                                  hjust = 0.5, size = 14), 
        plot.subtitle = element_text(size = 15))+
  labs(title = "Percent of population receiveing SNAP Benefits by US State and Region", 
       subtitle = str_wrap("Each gray line shows a state in a US region while the blue line shows the average for each region.\n 
                           Relative to other regions, people in the south are more likely to receive SNAP benefits. Despite near record high numbers of people that receive SNAP benefits, the percentage of the population that receive benefits has been on the decline since about 2014.", 
                           width = 95), 
       caption = "Source: U.S. Department of Agriculture")


fig2



```
:::

## Notes

Data was sourced from [Harvard Dataverse](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/AMUKJ0). Ten year state and national population data used to calculate the percentage of the population that receives SNAP benefits was obtained from the [USDA](https://data.ers.usda.gov/reports.aspx?ID=17827). There were some irregularities in the data where a large jump or decline (\> 40%) in participation rates in a particular month is followed by a large drop-off or jump (\> 40%) in the next month. In cases where there were such large irregularities between two consecutive months, observations were set to missing. The code for the project is available on [github](https://github.com/mikyasduga/SNAP_data_viz).
